### 2022-11-30 chapter1/chapter1.tex
Added note about scaling of axis for components figure
Updated text with new refinement process
Removed references to CNN in main text
Removed CNN methods and replaced with new methods
Removed references and citations to HMMer
Removed supplementary figures associated with CNN
Corrected a few uses of divergent instead of diverged

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0017-005F.png}
\centering
\caption{\textbf{Selected connected components of hit graph and summary statistics.}
\textbf{(A-B)} Two distinct connected components of the hit graph. Edges are colored by the value of their bit score. \textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component. \textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
\label{fig:components}
\end{figure}

@@

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

@@

\subsection*{Alignment refinement}
An initial alignment was created for each orthologous group as part of the selection of representative sequences. Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which are not aligned to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, in our pipeline the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments. Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini. This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps. These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations. Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them. Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state. In this case, however, the aligned sequences were processed individually and not as aligned columns. As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@

\subsection*{Alignment refinement}
Consensus columns were manually defined as columns in the initial alignments with at least five non-gap symbols using the Stockholm file format. These labeled alignments were used to build a profile HMM using HMMer 3.3.2~\cite{Eddy2009}. No sequence relative weighting was used, and the effective number of sequences was set to three times the number of sequences in the alignment. HMMer uses strong priors to allow the construction of profiles with relatively few aligned sequences. However, in rare cases a sequence with a long poorly supported segment was misaligned, effectively putting it out of register with the profile and creating a poor-quality alignment. Increasing the effective sequence number corrected this error. The original sequences were then aligned to the profile. Unaligned segments at the N- or C- terminus were removed from the alignment. Unaligned segments in internal regions were extracted and aligned separately for each region using MAFFT with the same settings. The resulting alignments were then “stitched” into the profile alignment. This process was unsuccessful for one alignment due to cryptic error in MAFFT, which reduced the number of refined alignments to 8,565.

This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences. However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM. A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM. The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output. The first layer was composed of four filters. Two of these filters had a kernel size of 10 and a dilation of 1. The other two filters had a kernel size of 60 with a dilation of 3. The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}). The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments. These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).

The trained network outputs a probability that each position in an aligned sequence is part of a long poorly supported segment. A simple algorithm was used to detect positions associated with high confidence long poorly supported segments even if the output at that position was relatively low. First, “seeds” were identified at positions where the output exceeded 0.75. These seeds were expanded outwards until the output was below 0.05. A seed and its expansion composed a long poorly supported segment. Consensus columns were defined as columns with at least five non-gap symbols, excluding positions labeled as long poorly supported segments. To maintain the contribution of long poorly supported segments to the profile’s models of insertions, long poorly supported segments were replaced with gaps only at positions aligned to consensus columns.

@@

\subsection*{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined with HMMer before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).

>>>>

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0017-005F.png}
\centering
\caption{\textbf{Selected connected components of hit graph and summary statistics.}
\textbf{(A-B)} Two distinct connected components of the hit graph. Edges are colored by the value of their bit score. \textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component. \textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node. Only the lower 99th percentile of the distribution is shown.}
\label{fig:components}
\end{figure}

@@

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section. Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.

@@

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, diverged regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Though the phylo-HMM removed phylogenetically incongruent ``insertions,'' some sequences still contained extensive segments of uninterrupted gaps. These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations. Unfortunately, they can also span more diverged regions, which complicates a simple rule- or threshold-based approach for identifying them. Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state. In this case, however, the aligned sequences were processed individually and not as aligned columns. As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. Since sequences produced from automated annotation pipelines can contain long segments which are not homologous to most or any other sequences in their respective orthologous groups, their alignments may contain over-aligned or poorly supported regions which can introduce artifacts into downstream analyses. Thus, in refinement over-alignment is avoided by aligning the sequences in two stages. In the first highly conserved regions are aligned using a strong correction for over-alignment, and in the second more diverged regions are aligned with a weaker correction. This process usually prevents errors caused by long poorly supported segments without degrading the quality of the alignment. In trimming, a phylo-HMM is used to remove regions which are poorly supported by the phylogenetic consensus.

While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@

\subsection*{Alignment refinement}
In the first stage, the representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section, except \textit{a\textsubscript{max}} was set to 0.7. A binary profile was created from the alignment using Gaussian process sequence weighting as described in the previous section. (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5. The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10. Diverged regions were taken as the complement of the highly conserved regions. For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4. The resulting sub-alignments were then stitched into the initial alignment.

@@

\subsection*{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).

### 2022-11-30 chapter1/chapter1_supporting.tex
Removed supporting figure renaming commands from tex file
Updated text with new refinement process
Removed references to CNN in main text
Removed CNN methods and replaced with new methods
Removed references and citations to HMMer
Removed supplementary figures associated with CNN
Corrected a few uses of divergent instead of diverged

\section*{Supporting information}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_architecture.png}
\centering
\caption{\textbf{CNN architecture.}}
\label{sfig:cnn_architecture}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
\centering
\caption{\textbf{CNN data and training details.}
\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment. \textbf{(B, C)} The training loss and metrics stabilized by the final training epochs. \textbf{(D, E)} The parameters for a model of this size are easily visualized. The embedding learns the gap and unknown amino acid symbols are distinct from the others. The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment. \textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN. The blue lines are the CNN’s predictions, and the yellow lines are the labels. The red lines are the weights given to each position. The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
\label{sfig:cnn_training}
\end{figure}

>>>>

\section*{Supporting information}

@@
